---
output:
  rmarkdown::github_document
always_allow_html: true
--- 
Homework 8

McKayla Hagerty 

CS 625, Fall 2020 

## Preparing Data

Original Source: https://data.world/kcmillersean/billboard-hot-100-1958-2017 (the orginial dataset was scraped using import.io)

Original Source: https://data.world/kcmillersean/billboard-hot-100-1958-2017 (the original dataset was pulled from Spotify Web API)

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



```{r}
#Import Billboard data.
billboard <- read.csv("https://query.data.world/s/j53mutis3fjnygz4fjctj576eljh32", header=TRUE, stringsAsFactors=FALSE)
billboard$WeekID <- as.Date(billboard$WeekID , format = "%m/%d/%Y")
View(billboard)
```




```{r}
library("httr")
library("readxl")
GET("https://query.data.world/s/5phlvfg3im7qyffzy66v4m2uvzmnr3", write_disk(tf <- tempfile(fileext = ".xlsx")))
genres_first_look <- read_excel(tf)
names(genres_first_look)
```



```{r}
#Cleaned genre data.
genres <- read.csv("C:/Users/13mck/Desktop/genres-final.csv")
names(genres)
head(genres)
```




```{r}
#Removing duplicate SongID rows in the genre dataset to allow for merging. 
genres <- genres[!duplicated(genres[c('SongID')]),] 
```


```{r}
merged<-merge(billboard, genres, by = "SongID",all.x=TRUE)
merged$WeekID <- as.Date(merged$WeekID , format = "%m/%d/%Y")

View(merged) 
```










```{r}
#Creating new column for year, month, and day. 
library(tidyr)

merged <- merged %>%
  separate(WeekID, sep="-", into = c("year", "month", "day"))
```


```{r}
#New dataframe for mean for each year.
library(dplyr)
mean_data <- group_by(merged, year) %>%
             summarise(danceability = mean(danceability, na.rm = TRUE),energy = mean(energy, na.rm = TRUE),speechiness = mean(speechiness, na.rm = TRUE),acousticness = mean(acousticness, na.rm = TRUE),liveness = mean(liveness, na.rm = TRUE),valence = mean(valence, na.rm = TRUE),loudness = mean(loudness, na.rm = TRUE))

mean_data <- as.data.frame(mean_data)

class(mean_data)

```

```{r}
library(ggplot2)
ggplot(data = mean_data, aes(x = year,group = 1)) + 
  geom_line(aes(y = danceability),color="blue") + 
  geom_line(aes(y = energy),color="green") +
  geom_line(aes(y = speechiness),color="purple") +
  geom_line(aes(y = acousticness),color="red") +
  geom_line(aes(y = liveness),color="orange") +
  geom_line(aes(y = valence)) +
  labs(y = "") # Delete or change y axis title if desired.
```

This is interesting. I can see how traits of what makes a top 100 song have changed over the years. This can especially be seen in the acousticness and valence. From previous findings, we would expect something interesting from the loudness as well. 

I also remember reading about radar charts and wonder if I could make an interactive radar chart with a slider to look into the traits for each year or decade. Or perhaps I could make one with an overlapping view with averages by decade. The problem with radar charts is that it is difficult to compare or get a sense of a specific measurement. However, if I layered them then comparing would be easier. Also, exact values are not the point of showing this data. The value is in the comparison between years or decades. 

The last trait I haven't spent much time on is the genre. I would like to see the change over time in what genres make the top 100 list. I'll first start with a stacked bar chart as use the years column rather than the WeekID column. *Has the popularity of different genres changed over time?*

```{r}
genrebyyear <- table(merged$year, merged$Genre) 
genrebyyear <- as.data.frame(genrebyyear)
names(genrebyyear)[names(genrebyyear) == "Var1"] <- "year"
names(genrebyyear)[names(genrebyyear) == "Var2"] <- "Genre"
names(genrebyyear)[names(genrebyyear) == "Freq"] <- "Count"
```

```{r}
# library
library(ggplot2)
topgenrebyyear <- genrebyyear %>% filter(Genre %in% c("rock", "pop","alternative/indie","electronic/dance","soul","hip-hop/rnb","country"))
  

# Stacked
ggplot(topgenrebyyear, aes(fill=Genre, y=Count, x=year)) + 
    geom_bar(position="stack", stat="identity")
```

I don't generally like stacked bar graphs because it is difficult to make comparisons between groups. However, I discovered streamgraphs which may be a better fit.

“Streamgraphs are a generalization of stacked area graphs where the baseline is free. By shifting the baseline, it is possible to minimize the change in slope (or wiggle) in individual series, thereby making it easier to perceive the thickness of any given layer across the data. Byron & Wattenberg describe several streamgraph algorithms in ‘Stacked Graphs—Geometry & Aesthetics3’”4

```{r}
#installing and loading streamgraph
#install.packages("devtools")
library(devtools)
#install.packages("zoo")
devtools::install_github("hrbrmstr/streamgraph")
topgenrebyyear$year <- as.Date(topgenrebyyear$year, format="%Y")
```

```{r}
library(dplyr)
library(streamgraph)

#topgenrebyyear %>% 
  #streamgraph(key="Genre", value="Count", date="year", interpolate="linear", sort=TRUE)%>%     #sg_fill_brewer("PuOr")%>%
  #sg_fill_manual(c("black", "#ffa500", "blue", "white", "#00ff00", "red")) %>% 
  #sg_axis_x(tick_interval=3, tick_units="year","%Y") 
  #sg_legend(FALSE, "Genre: ")
```

Should I choose to move forward with this chart, I would like to add key genre labels and more appropriate colors. The axis labels would also need adjusted. As mentioned before, if I continue down this route for my final visualization, I will probably return to the cleaning to see if I can show more of a representation of the pre-1990 song genres. 

At this point, Professor Weigle suggested I consider also looking into how particular genres have changed over time. To see if there was something interesting to be discovered there, I will create three new datasets with just rock, pop, and country music and create a multiple line plot for each featuring the music characteristics as before. *How have popular songs in key genres changed over time?* 

```{r}
library(ggplot2)
rockmusic <- merged %>% filter(Genre %in% c("rock"))
```

```{r}
#New dataframe for mean for each year for just rock music.
library(dplyr)
mean_data_rockmusic <- group_by(rockmusic, year) %>%
             summarise(danceability = mean(danceability, na.rm = TRUE),energy = mean(energy, na.rm = TRUE),speechiness = mean(speechiness, na.rm = TRUE),acousticness = mean(acousticness, na.rm = TRUE),liveness = mean(liveness, na.rm = TRUE),valence = mean(valence, na.rm = TRUE),loudness = mean(loudness, na.rm = TRUE))

mean_data_rockmusic <- as.data.frame(mean_data_rockmusic)

class(mean_data_rockmusic)

```


```{r}
p1 <- ggplot(data = mean_data_rockmusic, aes(x = year,group = 1)) + 
  geom_line(aes(y = danceability),color="blue") + 
  geom_line(aes(y = energy),color="green") +
  geom_line(aes(y = speechiness),color="purple") +
  geom_line(aes(y = acousticness),color="red") +
  geom_line(aes(y = liveness),color="orange") +
  geom_line(aes(y = valence)) +
  labs(y = "") # Delete or change y axis title if desired.
p1+theme(axis.text.x = element_text(angle = 90,vjust = .5))
```

This is interesting. I can see how rock music has changed since 1958.

```{r}
library(ggplot2)
popmusic <- merged %>% filter(Genre %in% c("pop"))
```

```{r}
#New dataframe for mean for each year for just pop music.
library(dplyr)
mean_data_popmusic <- group_by(popmusic, year) %>%
             summarise(danceability = mean(danceability, na.rm = TRUE),energy = mean(energy, na.rm = TRUE),speechiness = mean(speechiness, na.rm = TRUE),acousticness = mean(acousticness, na.rm = TRUE),liveness = mean(liveness, na.rm = TRUE),valence = mean(valence, na.rm = TRUE),loudness = mean(loudness, na.rm = TRUE))

mean_data_popmusic <- as.data.frame(mean_data_popmusic)

class(mean_data_popmusic)

```


```{r}
p1 <- ggplot(data = mean_data_popmusic, aes(x = year,group = 1)) + 
  geom_line(aes(y = danceability),color="blue") + 
  geom_line(aes(y = energy),color="green") +
  geom_line(aes(y = speechiness),color="purple") +
  geom_line(aes(y = acousticness),color="red") +
  geom_line(aes(y = liveness),color="orange") +
  geom_line(aes(y = valence)) +
  labs(y = "") # Delete or change y axis title if desired.
p1+theme(axis.text.x = element_text(angle = 90,vjust = .5))
```

```{r}
library(ggplot2)
countrymusic <- merged %>% filter(Genre %in% c("country"))
```

```{r}
#New dataframe for mean for each year for just country music.
library(dplyr)
mean_data_countrymusic <- group_by(countrymusic, year) %>%
             summarise(danceability = mean(danceability, na.rm = TRUE),energy = mean(energy, na.rm = TRUE),speechiness = mean(speechiness, na.rm = TRUE),acousticness = mean(acousticness, na.rm = TRUE),liveness = mean(liveness, na.rm = TRUE),valence = mean(valence, na.rm = TRUE),loudness = mean(loudness, na.rm = TRUE))

mean_data_countrymusic <- as.data.frame(mean_data_countrymusic)

class(mean_data_countrymusic)

```


```{r}
p1 <- ggplot(data = mean_data_countrymusic, aes(x = year,group = 1)) + 
  geom_line(aes(y = danceability),color="blue") + 
  geom_line(aes(y = energy),color="green") +
  geom_line(aes(y = speechiness),color="purple") +
  geom_line(aes(y = acousticness),color="red") +
  geom_line(aes(y = liveness),color="orange") +
  geom_line(aes(y = valence)) +
  labs(y = "") # Delete or change y axis title if desired.
p1+theme(axis.text.x = element_text(angle = 90,vjust = .5))
```

There is certainly much to work with here. There are interesting changes in all three genres.

## Final Questions: Results of the EDA

Question 1: *How has the prevalence of genres on the top 100 songs billboard changed over time?*

Sketch:

```{r}
#topgenrebyyear %>% 
  #streamgraph(key="Genre", value="Count", date="year", interpolate="linear", sort=TRUE)%>%     #sg_fill_brewer("PuOr")%>%
  #sg_fill_manual(c("black", "#ffa500", "blue", "white", "#00ff00", "red")) %>% 
  #sg_axis_x(tick_interval=3, tick_units="year","%Y") 
  #sg_legend(FALSE, "Genre: ")
```

The streamplot shows the main genre makeup of the top 100 songs from 1959 through 2019. It is easy to see how rock has steadily become less popular since 1986 as pop and hip hop have taken larger percentage of the billboard. I can also be seen that the soul genre has become less recognized since its surge of popularity in 1965 to 1980. Country, however, has become more popular in the 2000s.  


Question 2: *How has the pop, rock, and country genres changed over time? *

Sketches: (I will most likely choose this question for my final visualization.)

```{r}
p1 <- ggplot(data = mean_data_rockmusic, aes(x = year,group = 1)) + 
  geom_line(aes(y = danceability),color="blue") + 
  geom_line(aes(y = energy),color="green") +
  geom_line(aes(y = speechiness),color="purple") +
  geom_line(aes(y = acousticness),color="red") +
  geom_line(aes(y = liveness),color="orange") +
  geom_line(aes(y = valence)) +
  labs(y = "") # Delete or change y axis title if desired.
p1+theme(axis.text.x = element_text(angle = 90,vjust = .5))
```

```{r}
p1 <- ggplot(data = mean_data_popmusic, aes(x = year,group = 1)) + 
  geom_line(aes(y = danceability),color="blue") + 
  geom_line(aes(y = energy),color="green") +
  geom_line(aes(y = speechiness),color="purple") +
  geom_line(aes(y = acousticness),color="red") +
  geom_line(aes(y = liveness),color="orange") +
  geom_line(aes(y = valence)) +
  labs(y = "") # Delete or change y axis title if desired.
p1+theme(axis.text.x = element_text(angle = 90,vjust = .5))
```

```{r}
p1 <- ggplot(data = mean_data_countrymusic, aes(x = year,group = 1)) + 
  geom_line(aes(y = danceability),color="blue") + 
  geom_line(aes(y = energy),color="green") +
  geom_line(aes(y = speechiness),color="purple") +
  geom_line(aes(y = acousticness),color="red") +
  geom_line(aes(y = liveness),color="orange") +
  geom_line(aes(y = valence)) +
  labs(y = "") # Delete or change y axis title if desired.
p1+theme(axis.text.x = element_text(angle = 90,vjust = .5))
```



## Final Visualization

```{r}
library(ggplot2)
rockmusic <- merged %>% filter(merged$Genre %in% c("rock"))
```

```{r}
#New dataframe for mean for each year for just rock music.
library(dplyr)
rockmusic$Decade <- 10*as.integer(as.numeric(substring(as.character(rockmusic$year), 1,4)) / 10)
mean_data_rockmusic <- group_by(rockmusic, Decade) %>%
             summarise(danceability = mean(danceability, na.rm = TRUE),energy = mean(energy, na.rm = TRUE),speechiness = mean(speechiness, na.rm = TRUE),acousticness = mean(acousticness, na.rm = TRUE),liveness = mean(liveness, na.rm = TRUE),valence = mean(valence, na.rm = TRUE),loudness = mean(loudness, na.rm = TRUE))

mean_data_rockmusic <- as.data.frame(mean_data_rockmusic)

class(mean_data_rockmusic)

```

```{r}
library(ggplot2)
popmusic <- merged %>% filter(merged$Genre %in% c("pop"))
```

```{r}
#New dataframe for mean for each year for just pop music.
library(dplyr)
popmusic$Decade <- 10*as.integer(as.numeric(substring(as.character(popmusic$year), 1,4)) / 10)
mean_data_popmusic <- group_by(popmusic, Decade) %>%
             summarise(danceability = mean(danceability, na.rm = TRUE),energy = mean(energy, na.rm = TRUE),speechiness = mean(speechiness, na.rm = TRUE),acousticness = mean(acousticness, na.rm = TRUE),liveness = mean(liveness, na.rm = TRUE),valence = mean(valence, na.rm = TRUE),loudness = mean(loudness, na.rm = TRUE))

mean_data_popmusic <- as.data.frame(mean_data_popmusic)

class(mean_data_popmusic)

```

```{r}
library(ggplot2)
countrymusic <- merged %>% filter(merged$Genre %in% c("country"))
```

```{r warning=FALSE}
#New dataframe for mean for each year for just country music.
library(dplyr)
countrymusic$Decade <- 10*as.integer(as.numeric(substring(as.character(countrymusic$year), 1,4)) / 10)
mean_data_countrymusic <- group_by(countrymusic, Decade) %>%
             summarise(danceability = mean(danceability, na.rm = TRUE),energy = mean(energy, na.rm = TRUE),speechiness = mean(speechiness, na.rm = TRUE),acousticness = mean(acousticness, na.rm = TRUE),liveness = mean(liveness, na.rm = TRUE),valence = mean(valence, na.rm = TRUE),loudness = mean(loudness, na.rm = TRUE))

mean_data_countrymusic <- as.data.frame(mean_data_countrymusic)

class(mean_data_countrymusic)

```

```{r}
library(ggplot2)
	
hiphoprnbmusic <- merged %>% filter(merged$Genre %in% c("	
hip-hop/rnb"))
```

```{r warning=FALSE}
#New dataframe for mean for each year for just hiphop music.
library(dplyr)
hiphoprnbmusic$Decade <- 10*as.integer(as.numeric(substring(as.character(hiphoprnbmusic$year), 1,4)) / 10)
mean_data_hiphoprnbmusic <- group_by(countrymusic, Decade) %>%
             summarise(danceability = mean(danceability, na.rm = TRUE),energy = mean(energy, na.rm = TRUE),speechiness = mean(speechiness, na.rm = TRUE),acousticness = mean(acousticness, na.rm = TRUE),liveness = mean(liveness, na.rm = TRUE),valence = mean(valence, na.rm = TRUE),loudness = mean(loudness, na.rm = TRUE))

mean_data_hiphoprnbmusic <- as.data.frame(mean_data_hiphoprnbmusic)

class(mean_data_hiphoprnbmusic)

```

```{r fig.width=7, fig.height=6}

library(dplyr)
library(ggplot2)
library(scales)
library(grid)
library(cowplot)
library(gridExtra)

g1 <- ggplot(data = mean_data_countrymusic, aes(x = Decade,group = 1)) + 
  geom_line(size = 1.2,aes(y = danceability),color="light grey") + 
  geom_line(size = 1.2,aes(y = valence), color="dark grey")+
  geom_line(size = 2,aes(y = energy),color="mediumturquoise") +
  geom_line(size = 2,aes(y = acousticness),color="tomato2") +
  theme_bw() + 
  ylim(0, .9)+
  theme(legend.position = "none", axis.ticks = element_blank(), panel.border = element_blank(), panel.grid.major.x = element_blank(), panel.grid.minor.x = element_blank(), 
    axis.title.x = element_blank(),axis.text.y = element_blank(), axis.title.y = element_blank(), 
    axis.text.x = element_text( margin = margin(t = -8), size = 10),  
    plot.subtitle = element_text(face='bold.italic',size=10, margin = margin(b=0))) +
  annotate("text", x=2005, y=.63, label= "danceability",color="light grey", fontface =1,size = 3.5) +
  annotate("text", x=2005, y=.78, label= "energy",color="mediumturquoise", fontface =1,size = 3.5) +
  annotate("text", x=2005, y=.15, label= "acousticness",color="tomato2", fontface =1,size = 3.5) +
  annotate("text", x=2005, y=.47, label= "valence",color="dark grey", fontface =1,size = 3.5) +
  labs(subtitle = 'Qualities of Top Country Songs') 

g2 <- ggplot(data = mean_data_rockmusic, aes(x = Decade,group = 1)) + 
  geom_line(size = 1.2,aes(y = danceability),color="light grey") + 
  geom_line(size = 1.2,aes(y = valence), color="dark grey")+
  geom_line(size = 2,aes(y = energy),color="mediumturquoise") +
  geom_line(size = 2,aes(y = acousticness),color="tomato2") +
  theme_bw() + 
  ylim(0, .9)+
  theme(legend.position = "none", axis.ticks = element_blank(), panel.border = element_blank(), panel.grid.major.x = element_blank(), panel.grid.minor.x = element_blank(), 
    axis.title.x = element_blank(),axis.text.y = element_blank(), axis.title.y = element_blank(), 
    axis.text.x = element_text( margin = margin(t = -8), size = 10),  
    plot.subtitle = element_text(face='bold.italic',size=10, margin = margin(b=0))) +
  annotate("text", x=2005, y=.6, label= "danceability",color="light grey",fontface =1,size = 3.5) +
  annotate("text", x=2005, y=.83, label= "energy", color="mediumturquoise", fontface =1,size = 3.5) +
  annotate("text", x=2005, y=.14, label= "acousticness",color="tomato2",fontface =1,size = 3.5) +
  annotate("text", x=2005, y=.45, label= "valence",color="dark grey",fontface =1,size = 3.5) +
  labs( subtitle = 'Qualities of Top Rock Songs') 

g3 <- ggplot(data = mean_data_popmusic, aes(x = Decade,group = 1)) + 
  geom_line(size = 1.2,aes(y = danceability),color="light grey") + 
  geom_line(size = 1.2,aes(y = valence), color="dark grey")+
  geom_line(size = 2,aes(y = energy),color="mediumturquoise") +
  geom_line(size = 2,aes(y = acousticness),color="tomato2") +
  theme_bw() + 
  ylim(0, .9)+
  theme(legend.position = "none", axis.ticks = element_blank(), panel.border = element_blank(), panel.grid.major.x = element_blank(), panel.grid.minor.x = element_blank(), 
    axis.title.x = element_blank(),axis.text.y = element_blank(), axis.title.y = element_blank(), 
    axis.text.x = element_text( margin = margin(t = -8), size = 10),  
    plot.subtitle = element_text(face='bold.italic',size=10, margin = margin(b=0))) +
  annotate("text", x=2005, y=.625, label= "danceability",color="light grey",fontface =1,size = 3.5) +
  annotate("text", x=2005, y=.75, label= "energy",color="mediumturquoise",fontface =1,size = 3.5) +
  annotate("text", x=2005, y=.22, label= "acousticness",color="tomato2",fontface =1,size = 3.5) +
  annotate("text", x=2005, y=.51, label= "valence",color="dark grey",fontface =1,size = 3.5) +
  labs( subtitle = 'Qualities of Top Pop Songs') 

g4 <- ggplot(data = mean_data_hiphoprnbmusic, aes(x = Decade,group = 1)) + 
  geom_line(size = 1.2,aes(y = danceability),color="light grey") + 
  geom_line(size = 1.2,aes(y = valence), color="dark grey")+
  geom_line(size = 2,aes(y = energy),color="mediumturquoise") +
  geom_line(size = 2,aes(y = acousticness),color="tomato2") +
  theme_bw() + 
  ylim(0, .9)+
  theme(legend.position = "none", axis.ticks = element_blank(), panel.border = element_blank(), panel.grid.major.x = element_blank(), panel.grid.minor.x = element_blank(), 
    axis.title.x = element_blank(),axis.text.y = element_blank(), axis.title.y = element_blank(), 
    axis.text.x = element_text( margin = margin(t = -8), size = 10),  
    plot.subtitle = element_text(face='bold.italic',size=10, margin = margin(b=0))) +
  annotate("text", x=2005, y=.63, label= "danceability",color="light grey",fontface =1,size = 3.5) +
  annotate("text", x=2005, y=.78, label= "energy",color="mediumturquoise",fontface =1,size = 3.5) +
  annotate("text", x=2005, y=.15, label= "acousticness",color="tomato2",fontface =1,size = 3.5) +
  annotate("text", x=2005, y=.48, label= "valence",color="dark grey",fontface =1,size = 3.5) +
  labs( subtitle = 'Qualities of Top Hip-Hop/R&B Songs') 

grid.arrange(g1, g2, g3, g4, ncol=2,
     top = textGrob("Want to write a hit song? Maybe it's time to leave the acoustic guitar behind",x = 0, hjust = 0,gp=gpar(fontfamily="sans", fontsize=13.7,font=2,lineweight=1)))

```

## Explaination 






## References:

https://stackoverflow.com/questions/40663586/streamgraph-package-in-r-will-run-the-data-but-no-graph-will-be-shown
https://www.quora.com/How-do-I-get-a-frequency-count-based-on-two-columns-variables-in-an-R-dataframe
https://stackoverflow.com/questions/10758961/how-to-convert-a-table-to-a-data-frame
https://www.datanovia.com/en/lessons/rename-data-frame-columns-in-r/
https://www.datanovia.com/en/lessons/subset-data-frame-rows-in-r/
https://rdrr.io/github/hrbrmstr/streamgraph/man/sg_axis_x.html
https://www.r-graph-gallery.com/158-change-color-in-interactive-streamgraph.html
https://stackoverflow.com/questions/3171426/compare-two-data-frames-to-find-the-rows-in-data-frame-1-that-are-not-present-in
https://community.rstudio.com/t/converting-dates-year-month-day-to-3-separate-columns-year-month-day/8585/3
https://stackoverflow.com/questions/35920408/creating-a-line-chart-in-r-for-the-average-value-of-groups
https://stackoverflow.com/questions/30375600/how-to-plot-multiple-lines-for-each-column-of-a-data-matrix-against-one-column
https://stackoverflow.com/questions/27082601/ggplot2-line-chart-gives-geom-path-each-group-consist-of-only-one-observation
https://stackoverflow.com/questions/15215457/standardize-data-columns-in-r
https://www.earthdatascience.org/courses/earth-analytics/time-series-data/summarize-time-series-by-month-in-r/ 